---
title: "Doing Bayesian Data Analysis Ch 13"
output: 
  ioslides_presentation: 
    keep_md: yes
---

# Power Analysis


## Experiment goals

* Reject a null value
    + HDI excludes region of practical equivalence around null

* Accept a null value
    + HDI includes null

* Achieve precision in estimation
    + width of HDI is below a given maximum
    

## Experiments and power

* These goals can only be achieved probablistically.

* The probability of achieving the goal (e.g. rejecting a null value) is the 
power of the experiment.

* Methods of increasing power:
    + Reduce experiment noise (e.g. remove other random influences)
    + Increase magnitude of effect (e.g. use large doses of treatment)
    + Increase sample size


## Calculating power

* Depends on
    + hypothetical distribution of effects in the population being measured
    + a specified data-sampling plan


## Calculating power

1. From the hypothetical distribution of parameter values, randomly generate
representative values.
2. From the representative parameter values, generate a random sample of 
data, using the planned sampling method.
3. From the simulated sample of data, compute the posterior estimate.
4. From the posterior estimate, tally whether or not the goals were attained.
5. Repeat the above steps many times, to approximate the power.


## Example

model: y ~ normal(mu, sigma)

null: mu = 0

find the sample size necessary to reject null with probability p


## Example: model definition {.smaller}

```{r, message=FALSE, warning=FALSE, results='hide'}
library(rstan)

model_code <- "
data {
  int<lower=0> N;
  real y[N];
}
parameters {
  real mu;
  real<lower=0> sigma;
}
model {
  y ~ normal(mu, sigma);
}
"

# build the model; do not estimate
fit <- stan(model_code = model_code, 
            data = list(N = 2, y = c(0, 1)), 
            chains = 1, iter = 1)
```


## Example: power calculation {.smaller}

```{r, echo=FALSE}
hdi <- function(sampleVec, credMass=0.95) {
  # Computes highest density interval from a sample of representative values,
  #   estimated as shortest credible interval.
  # Arguments:
  #   sampleVec
  #     is a vector of representative values from a probability distribution.
  #   credMass
  #     is a scalar between 0 and 1, indicating the mass within the credible
  #     interval that is to be estimated.
  # Value:
  #   HDIlim is a vector containing the limits of the HDI
  sortedPts = sort( sampleVec )
  ciIdxInc = ceiling( credMass * length( sortedPts ) )
  nCIs = length( sortedPts ) - ciIdxInc
  ciWidth = rep( 0 , nCIs )
  for ( i in 1:nCIs ) {
    ciWidth[ i ] = sortedPts[ i + ciIdxInc ] - sortedPts[ i ]
  }
  HDImin = sortedPts[ which.min( ciWidth ) ]
  HDImax = sortedPts[ which.min( ciWidth ) + ciIdxInc ]
  HDIlim = c( HDImin , HDImax )
  return( HDIlim )
}
```

```{r}
sample_sizes <- c(50, 100, 200)
iter <- 1000
sim_results <- matrix(nrow = length(sample_sizes), ncol = iter)
rownames(sim_results) <- paste0("n", sample_sizes)

overlap <- function(x, y) {x[1] <= y[2] & y[1] <= x[2]}

for (i in seq_along(sample_sizes)) {
  for (j in seq(iter)) {
    sample_size <- sample_sizes[i]
    
    # 1. generate representative value
    mu <- rnorm(1)
    
    # 2. generate random sample
    y <- rnorm(sample_size, mean = mu)
    data <- list(N = sample_size, y = y)
    
    # 3. estimate posterior
    posterior <- stan(data = data, fit = fit, chains = 1)
    
    # 4. does HDI exclude ROPE?
    hdi_mu <- hdi(extract(posterior)$mu)
    success <- !overlap(c(-0.02, 0.02), hdi_mu)
    sim_results[i, j] <- success
  }
}
```


## Example: results

model: y ~ normal(mu, sigma)

null: mu = 0

power: probability of rejecting the null, by sample size

```{r}
apply(sim_results, 1, mean)
```
